---
title: "Dynamic Linear Models and Cholera in the DRC"
author: "Jacob Carey"
date: "May 14, 2015"
bibliography: references.bib
output: 
  beamer_presentation:
    theme: "Szeged"
    toc: true
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r libs}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
theme_set(theme_classic())
```

```{r data}
source("~/Code/Dynamics/cholera/scripts/read-in.R")
input.dir <- "~/Code/Dynamics/epicentrehopkins/trunk/Data/DRC/kalemie"
data <- read.in(input.dir) %>%
    tbl_df() %>%
    filter(!AS %in% c("", "Out of Area", "MAHILA"),
           !is.na(Date),
           year(Date) > 2008)
```

# Background {-}

- Cholera is endemic in the DRC
- Vaccine exists
- Vaccine Campaign conducted by MSF

# Overview
```{r}
ggplot(data, aes(x=factor(year))) +
    geom_histogram() +
    xlab("Year") +
    ylab("Cholera Cases")
```

# Overview
```{r}
graph.data <- data %>%
    group_by(region=AS, quarter=quarter(Date, with_year=TRUE)) %>%
    summarise(cases=n()) %>%
    mutate(quarter.dec=(quarter - floor(quarter) - 0.1) * 2.5,
           quarter=floor(quarter) + quarter.dec) %>%
    select(-quarter.dec)

ggplot(graph.data, aes(x=quarter, y=cases)) +
    geom_point(aes(colour=region)) +
    geom_smooth(aes(colour=region), se=FALSE) +
    xlab("Quarter") +
    ylab("Cholera Cases")
```

# Dynamic Linear Models

- Special case of State Space Models.

- Described in @dlm.

# Methods

- Fit DLM to data

# Results

- More graphs

# Discussion

- Vaccine resulted in 20% fewer cases in vaccinated regions

# Future work

- Nonparametric Bayesian statistics

# References

---

## Questions?

