---
title: "Dynamic Linear Models and Cholera in the DRC"
author: "Jacob Carey"
date: "May 14, 2015"
bibliography: references.bib
header-includes:
   - \usepackage{tikzsymbols}
output: 
  beamer_presentation:
    theme: "Szeged"
    slide_level: 2
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r libs}
library(dlm)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
theme_set(theme_classic())
```

```{r funcs}
source("./../R/dlmFilterDF.R")
source("./../R/model.R")
```

```{r data}
source("~/Code/Dynamics/cholera/scripts/read-in.R")
input.dir <- "~/Code/Dynamics/epicentrehopkins/trunk/Data/DRC/kalemie"
data <- read.in(input.dir) %>%
    tbl_df() %>%
    filter(!AS %in% c("", "Out of Area", "MAHILA"),
           !is.na(Date),
           year(Date) > 2009)

cases <- data %>%
    group_by(date=round_date(Date, unit="week")) %>%
    summarise(cases=n())
```

# Background
## Cholera

- Infectious disease presenting with diarrhea
- Often spread through contaminated water
- Treatment is simple and cheap
    - Though time consuming
- But vaccine exists (not licensed in US)  
  
@cholera

## My problem

- Cholera is endemic in the DRC
- Vaccine Campaign conducted by MSF
- Hard to implement campaign

## Question

- Vaccine campaign was only conducted in some of the areas
- Was the campaign effective? If so, how effective was it?

# Summary of data
## Cholera Cases by Year
```{r}
ggplot(data, aes(x=factor(year))) +
    geom_histogram() +
    xlab("Year") +
    ylab("Cholera Cases")
```

## Cholera by Region and Time
```{r}
graph.data <- data %>%
    group_by(region=AS, quarter=quarter(Date, with_year=TRUE)) %>%
    summarise(cases=n()) %>%
    mutate(quarter.dec=(quarter - floor(quarter) - 0.1) * 2.5,
           quarter=floor(quarter) + quarter.dec) %>%
    select(-quarter.dec)

ggplot(graph.data, aes(x=quarter, y=cases)) +
    geom_point(aes(colour=region)) +
    geom_smooth(aes(colour=region), se=FALSE) +
    xlab("Quarter") +
    ylab("Cholera Cases")
```

## Univariate Time Series for Analysis
```{r}
ggplot(cases, aes(x=date, y=cases)) +
    geom_line() +
    geom_point()
```

# Analytical Considerations
## Challenges

- Temporal trend
    - Assumption of equal variance doesn't hold
- Infectious disease
    - Assumption of independence doesn't hold

## Can These Work?

- GLM
    - Not all assumptions hold
- Changepoint analysis
    - Doesn't account for temporal/seasonal trends

## Dynamic Linear Models?

- Dynamic Linear Models
- Special case of State Space Models.
- Described in @dlm.

# Dynamic Linear Models
## Formulation of Time Series

- Time series formulation
- $\theta_t = \theta_{t-1} + \nu + w_t$
    - $w_t \sim \text{Normal}(0, \sigma_w^2)$  
- $Y_t = \theta_t + \epsilon_t$
    - $\epsilon_t \sim \text{Normal}(0, \sigma^2)$

## Formulation of Dynamic Linear Models

- Dynamic Linear Models specification
- $\theta_{t=0} \sim \text{Normal}(m_0, C_0)$
- $Y_t = F_t \theta_t + v_t$
    - $v_t \sim \text{Normal}(0, V_t)$
- $\theta_t = G_t \theta_{t-1} + w_t$
    - $w_t \sim \text{Normal}(0, W_t)$

## What about Bayes?

- Those unknowns are intimidating...
- MLE can be used, but often times many local maxima
- Sometimes a relatively flat likelihood
- Let's use Bayesian estimation instead

## Some Notes about DLMs

- Naturally extends to multivariate time series
    - Covariance between regional cases makes spatial sense
- Online forecasting
    - Not really relevant to my problem but an interesting 
      finding.
- DLMs can also be used in combination with regression.
    - Not considered here

# Findings
## Results

```{r results, warning=FALSE}
y <- ts(cases$cases, freq=52)
dates <- cases$date
data.model <- data.frame(dlm.model(y, order=4)) %>%
    mutate(x=dates)

data.model <- data.model[10:nrow(data.model), ]

ggplot(data.model, aes(x=x, y=obs)) +
    geom_point() +
    geom_ribbon(aes(ymin=ll, ymax=ul),
                alpha=0.25, colour="blue") +
    xlab("Date") + 
    ylab("Cases") +
    ggtitle("One-ahead Forecast")
```

## Discussion

- Did not see evidence of vaccine effectiveness $\Sadey$
- This could likely be due to level of detail lost in aggregating regions
- More work needed to clarify this point
    - Possible that there was no overall change in incidence of Cholera
    - Concerning! $\NiceReapey$

## Future work

- Multivariate case
    - I've structured the data as a univariate time series for simplification
    - There's definitely covariance between regions that needs to be accounted
      for
- Weather likely plays a role in this data
    - Next steps should include collecting weather data to use in a 
      DLM regression.

---

> - Questions?
> - Or should we just call it summer?

## References

