---
title:  Homework 1
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

**Libraries**
```{r libraries}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
# library(MASS)
library(mvtnorm)
library(ggplot2)
set.seed(1)
```

**Problem 1**  
**a)**
We sample from a multivariate normal with given prior parameters.
The samples are "jittered" for inspection.  
```{r p1a}
a.0 <- -17.31
b.0 <- 2.57
Sigma.0 <- matrix(c(1053.72, -156.45, -156.45, 23.24), nrow=2)

prior.samples <- rmvnorm(1e3, c(a.0, b.0), Sigma.0)

ggplot(data.frame(prior.samples), aes(x=X1, y=X2)) +
   geom_jitter() +
   xlab(expression(alpha)) +
   ylab(expression(beta))
```

**b)**
We transform the log dose to log mg/ml before the following problems.  
We create a grid of $\alpha$ from -80 to -5 and $\beta$ from 0 to 11.
Each $\alpha$, $\beta$ combination from the grid is used to calculate a
probability of death via the standard normal CDF of 
$\alpha + \beta \times x_i$ for each $i = 1, ..., 5$. The log likelihood
is computed as the sum of $y_i \times \log(p_i) + (5 - y_i) \times \log(1 - p_i)$.  

```{r p1b}
x <- c(-0.86, -0.30, -0.05, 0.73) + log(1e3)
n <- rep(5, 4)
y <- c(0, 1, 3, 5)

lik <- function(grid, x, y, n) {
    a <- grid[1]
    b <- grid[2]
    p <- pnorm(a + b * x)
    log.lik <- sum(y * log(p) + (n - y) * log(1 - p))
    return(exp(log.lik))
}

a.grid <- seq(-80, -5, 0.1)
b.grid <- seq(0, 11, 0.1)
grid <- expand.grid(a.grid, b.grid)

z <- matrix(apply(grid, 1, lik, x=x, y=y, n=n),
            nrow=length(a.grid), ncol=length(b.grid),
            byrow=FALSE)

z[is.na(z)] <- 0

contour(a.grid, b.grid, z, 
        xlab=expression(alpha), 
        ylab=expression(beta), drawlabels=FALSE,
        main="Likelihood Plot")
```

The likelihood has a similar shape to the prior, however, it
appears to be more concentrated than the prior.

**c)**
Using the same grid from problem 1b, we calculate the PDF of the
multivariate normal for each combination. This PDF is multiplied
with the likelihood calculated in 1b to obtain the generalized
likelihood.  

```{r p1c}
gen.lik <- function(grid, x, y, n) {
    lik <- lik(grid, x, y, n)
    prior.prob <- dmvnorm(grid, c(a.0, b.0), Sigma.0)
    return(lik * prior.prob)
}

z <- matrix(apply(grid, 1, gen.lik, x=x, y=y, n=n),
            nrow=length(a.grid), ncol=length(b.grid),
            byrow=FALSE)

contour(a.grid, b.grid, z, 
        xlab=expression(alpha), 
        ylab=expression(beta), drawlabels=FALSE,
        main="Generalized Likelihood Plot")
```

**d)**
We calculate the joint posterior density using the
by normalizing the generalized likelihood from 1c.
The marginal probabilities are calculated by 
summing over $\alpha$ and $\beta$ for $\beta$
and $\alpha$ respectively. We sample from the
joint distribution as described in lecture 2.  
```{r p1d}
GenRandomFromGrid <- function(x, cdf.x) {
    u <- runif(1)
    max(x[u > cdf.x])
}

p.ab <- z/sum(z, na.rm=TRUE)
p.a <- apply(p.ab, 1, sum, na.rm=TRUE)
p.b <- apply(p.ab, 2, sum, na.rm=TRUE)

cdf.a <- cumsum(p.a)
cdf.b <- cumsum(p.b)

u <- runif(1e3)

a <- rep(NA, length(u))
b <- rep(NA, length(u))

for (i in 1:length(u)) {
    a[i] <- max(a.grid[u[i] > cdf.a])
}

for (i in 1:length(a)) {
    j <- which(a.grid == a[i])
    cdf.b.given.a <- cumsum(na.omit(p.ab[j, ])) / 
        sum(p.ab[j, ], na.rm=TRUE)
    b[i] <- max(b.grid[u[i] > cdf.b.given.a])
}

plot(a, b, xlab=expression(alpha), ylab=expression(beta))
```

**e)**
We use the marginal posteriors calculated in 1d.  
```{r p1e}
qplot(a, xlab=expression(alpha), binwidth=3)
qplot(b, xlab=expression(beta), binwidth=0.1)
```

**f)**
We use our samples from the joint posterior
distribution to calculate the posterior density of the
log dose that is lethal to 50% of the animal.  
```{r p1f}
log.ld.50 <- sort(-a / b)

qplot(log.ld.50, xlab=expression("Log LD"[50]), 
      ylab="Density", binwidth=0.3)
```

**g)**
Since the logarithm is a continuous and increasing function with inverse
$\exp$, we use the change of variable technique to obtain the 
posterior density of the $\text{LD}_{50}$. We do this by computing
the CDF of the log LD50 (cumsum(log ld50)/sum(log ld 50)). This is
related to the CDF of the LD50 by taking antilog of the concentrations
estimated in log ld50 scale.
```{r p1g}
# P(log ld_50 < log(x))
cdf <- cumsum(log.ld.50) / sum(log.ld.50)
ld.50 <- exp(log.ld.50)
qplot(ld.50, cdf)
# qplot(log.ld.50, log.ld.50.cdf)
ld.50 <- log.ld.50 / abs(exp(log.ld.50))
qplot(ld.50, xlab=expression("LD"[50]), ylab="Density")
```

The posterior mean is `r mean(ld.50)`, median: `r median(ld.50)`,
standard deviation `r sd(ld.50)`, 90% credible interval
`r quantile(ld.50, c(0.05, 0.95))`
