---
title:  Homework 4
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE,
                      cache=FALSE)
```

```{r libraries}
library(dplyr)
library(readr)
library(ggplot2)
library(R2jags)
```

```{r data}
fungus <- read_delim("./ToenailFungusData.txt", delim=" ")
```

# Problem 1
Write out a full model (sampling distribution and priors) to characterize these repeated binary data, using a logit link function. Assume that the only subject-apecific parameter is the intercept term. (Note: Since this is a randomized study, the average score for patients on each treatment should be the same at week 0. Therefore, if there is a treatment effect, the treatment-specific slopes of the scores over time will differ.)

# Problem 2
Fit the model and summarize the results. What do you conclude about the treatment effect?

# Problem 3
Now write out a full Bayesian hierarchical model, allowing each patient to have his or her own regression coefficients.

# Problem 4
Fit the full hierarchical model and summarize the results. Has your inference about the relative effectiveness of the treatments changed?
