---
title:  Homework 4
author: "[Jacob Carey](mailto:jcarey15@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      cache=FALSE,
                      results='asis')
```

```{r libraries}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(R2jags)
theme_set(theme_classic())
```

```{r data}
fungus <- read_delim("./ToenailFungusData.txt", delim=" ") %>%
    # reallocate visit number
    group_by(ID) %>%
    mutate(Visit=1:n())

y <- fungus %>% select(ID, Visit, y) %>% spread(Visit, y) %>% select(-ID)
treatment <- fungus %>% distinct(ID, Trt) %>% select(ID, Trt) %>% .$Trt
time <- fungus %>% select(ID, Visit, Time) %>% spread(Visit, Time) %>%
    select(-ID)
n.visits <- fungus %>% count(ID) %>% .$n
n.people <- length(n.visits)
```

# Problem 1
Write out a full model (sampling distribution and priors) to characterize these repeated binary data, using a logit link function. Assume that the only subject-apecific parameter is the intercept term. (Note: Since this is a randomized study, the average score for patients on each treatment should be the same at week 0. Therefore, if there is a treatment effect, the treatment-specific slopes of the scores over time will differ.)

$$
\begin{aligned}
y_{i,j} | x_i, t_{i, j}, \alpha, \alpha_{i}, 
\beta_1, \beta_2, \beta_3 \sim 
\text{bernoulli}(\mu_{i, j}) \\
\text{logit}(\mu_{i, j}) = \alpha + \alpha_{i} + 
\beta_1 x_i + \beta_2 t_{i, j} + \beta_3 x_i t_{i, j} \\
\alpha, \beta_1, \beta_2 \beta_3, \sim \text{Normal}(0, 1000) \\
\alpha_{i} \sim \text{Normal}(\mu_{\alpha}, \omega_{\alpha}^2) \\
\omega_{\alpha}^2 \sim \text{Unif(0, 100)}
\end{aligned}
$$

# Problem 2
Fit the model and summarize the results. What do you conclude about the treatment effect?

```{r p2}
fungus.model <- function() {
    for (i in 1:n.people) {
        for (j in 1:n.visits[i]) {
            y[i, j] ~ dbern(mu[i, j])
            logit(mu[i, j]) <- alpha.fixed + alpha[i] + 
                               (b1.fixed) * treatment[i] +
                               (b2.fixed) * time[i, j] +
                               (b3.fixed) * treatment[i] * time[i, j]
        }

        alpha[i] ~ dnorm(0.0, omega.a.inv) # random intercept
    }

    alpha.fixed ~ dnorm(0.0, 1.0e-04) # fixed intercept
    b1.fixed ~ dnorm(0.0, 1.0e-04) # fixed treatment coef
    b2.fixed ~ dnorm(0.0, 1.0e-04) # fixed time coef
    b3.fixed ~ dnorm(0.0, 1.0e-04) # fixed interaction coef
    omega.a.inv <- 1 / omega.a
    omega.a ~ dunif(0, 10) # between group variance
}

fungus.data <- list("y", "treatment", "time",
                    "n.people", "n.visits")
fungus.params <- c("alpha.fixed", "omega.a", 
                   "b1.fixed", "b2.fixed", 
                   "b3.fixed")
fungus.inits <- function() {
    list("alpha.fixed"=rnorm(1, 0, 1),
         "omega.a"=runif(1, 0, 1),
         "b1.fixed"=rnorm(1, 0, 1),
         "b2.fixed"=rnorm(1, 0, 1),
         "b3.fixed"=rnorm(1, 0, 1))
}

fungus.fit <- jags(data=fungus.data, inits=fungus.inits,
                   parameters.to.save=fungus.params,
                   model.file=fungus.model, n.chains=3, progress.bar="none",
                   n.iter=5e3)

output <- fungus.fit$BUGSoutput$summary
rownames(output) <- c("$\\alpha$", "$\\beta_1$", "$\\beta_2$", "$\\beta_3$",
                      "deviance", "$\\omega_{\\alpha}^2$")
knitr::kable(output, digits=3)
```


I conclude that the treatment is effective. Using this hierarchical model, the effectiveness of treatment seems small.

# Problem 3
Now write out a full Bayesian hierarchical model, allowing each patient to have his or her own regression coefficients.

$$
\begin{aligned}
y_{i,j} | x_i, t_{i, j}, \alpha, \alpha_{i}, 
\beta_1, \beta_2, \beta_3,
\beta_{1_i}, \beta_{2_i}, \beta_{3_i} \sim 
\text{bernoulli}(\mu_{i, j}) \\
\text{logit}(\mu_{i, j}) = (\alpha + \alpha_{i}) + 
(\beta_1 + \beta_{1_i}) x_i + 
(\beta_2 + \beta_{2_i}) t_{i, j} + 
(\beta_3 + \beta_{3_i}) x_i t_{i, j} \\
\alpha, \beta_1, \beta_2, \beta_3, \sim \text{Normal}(0, 1000) \\
\alpha_{i} \sim \text{Normal}(\mu_{\alpha}, \omega_{\alpha}^2) \\
\beta_{1_i} \sim \text{Normal}(\mu_{\beta_1}, \omega_{\beta_1}^2) \\
\beta_{2_i} \sim \text{Normal}(\mu_{\beta_2}, \omega_{\beta_2}^2) \\
\beta_{3_i} \sim \text{Normal}(\mu_{\beta_3}, \omega_{\beta_3}^2) \\
\omega_{\alpha}^2 \sim \text{Unif(0, 100)}
\end{aligned}
$$

# Problem 4
Fit the full hierarchical model and summarize the results. Has your inference about the relative effectiveness of the treatments changed?

```{r p4}
fungus.model <- function() {
    for (i in 1:n.people) {
        for (j in 1:n.visits[i]) {
            y[i, j] ~ dbern(mu[i, j])
            logit(mu[i, j]) <- alpha.fixed + alpha[i] + 
                               (b1.fixed + b1[i]) * treatment[i] +
                               (b2.fixed + b2[i]) * time[i, j] +
                               (b3.fixed + b3[i]) * treatment[i] * time[i, j]
        }

        alpha[i] ~ dnorm(0.0, omega.a.inv) # random intercept
        b1[i] ~ dnorm(0.0, omega.b1.inv) # random treatment
        b2[i] ~ dnorm(0.0, omega.b2.inv) # random time
        b3[i] ~ dnorm(0.0, omega.b3.inv) # random interaction
    }

    alpha.fixed ~ dnorm(0.0, 1.0e-04) # fixed intercept
    b1.fixed ~ dnorm(0.0, 1.0e-04) # fixed treatment coef
    b2.fixed ~ dnorm(0.0, 1.0e-04) # fixed time coef
    b3.fixed ~ dnorm(0.0, 1.0e-04) # fixed interaction coef
    omega.a.inv <- 1 / omega.a
    omega.b1.inv <- 1 / omega.b1
    omega.b2.inv <- 1 / omega.b2
    omega.b3.inv <- 1 / omega.b3
    omega.a ~ dunif(0, 10) # between group variance
    omega.b1 ~ dunif(0, 10) # between group variance
    omega.b2 ~ dunif(0, 10) # between group variance
    omega.b3 ~ dunif(0, 10) # between group variance
}

fungus.data <- list("y", "treatment", "time",
                    "n.people", "n.visits")
fungus.params <- c("alpha.fixed", "omega.a", 
                   "omega.b1", "omega.b2", "omega.b3",
                   "b1.fixed", "b2.fixed", 
                   "b3.fixed")
fungus.inits <- function() {
    list("alpha.fixed"=rnorm(1, 0, 1),
         "omega.a"=runif(1, 0, 1),
         "omega.b1"=runif(1, 0, 1),
         "omega.b2"=runif(1, 0, 1),
         "omega.b3"=runif(1, 0, 1),
         "b1.fixed"=rnorm(1, 0, 1),
         "b2.fixed"=rnorm(1, 0, 1),
         "b3.fixed"=rnorm(1, 0, 1))
}

fungus.fit <- jags(data=fungus.data, inits=fungus.inits,
                   parameters.to.save=fungus.params,
                   model.file=fungus.model, n.chains=3, progress.bar="none",
                   n.iter=5e4, n.burnin=5e3)

output <- fungus.fit$BUGSoutput$summary
rownames(output) <- c("$\\alpha$", "$\\beta_1$", "$\\beta_2$", "$\\beta_3$",
                      "deviance", "$\\omega_{\\alpha}^2$", 
                      "$\\omega_{\\beta_1}^2$", "$\\omega_{\\beta_2}^2$",
                      "$\\omega_{\\beta_3}^2$")
knitr::kable(output, digits=3)
```


My inference about the direction of effectiveness is still the same. However, when fitting a full hierarchical model, the population level effect of treatment is stronger - there is a larger protective effect of the treatment.
